---
title: "Milestone Project"
author: "Jaime Cappa"
date: "March 7, 2016"
output: html_document
---

#JHU Data Science Capstone Milestone Report

This report is produced in partial fulfillment of the requirements for the Capstone Project offered by Johns Hopkins Bloomberg School of Public Health and Coursera.

This report describes the exploratory data analysis of the Capstone Dataset.

```{r, echo = TRUE}
paste("Check file sizes in MB")
```

```{r, echo = TRUE}
file.info("en_US.twitter.txt")$size / (1024*1024)
```

```{r, echo = TRUE}
file.info("en_US.news.txt")$size / (1024*1024)
```

```{r, echo = TRUE}
file.info("en_US.blogs.txt")$size / (1024*1024)
```

###View Line Counts:

```{r, echo = TRUE}
install.packages('R.utils', repos='http://cran.us.r-project.org')
library(R.utils)
countLines("en_US.twitter.txt")
```

```{r, echo = TRUE}
countLines("en_US.news.txt")
```

```{r, echo = TRUE}
countLines("en_US.blogs.txt")
```

###View Word Counts:

```{r, echo = TRUE}
system2("wc", args = "-L en_US.twitter.txt", stdout=TRUE)
```

```{r, echo = TRUE}
system2("wc", args = "-L en_US.news.txt", stdout=TRUE)
```

```{r, echo = TRUE}
system2("wc", args = "-L en_US.blogs.txt", stdout=TRUE)
```

#Perform Sampling

Given the large amount of text and limited computational resources, sampling is performed. 10000 lines per file is randomly sampled and saved to disk.

```{r, echo = TRUE}
twitter <- readLines('en_US.twitter.txt')
news <- readLines('en_US.news.txt')
blogs <- readLines('en_US.blogs.txt')
set.seed(39)
sampleTwitter <- twitter[sample(1:length(twitter),10000)]
sampleNews <- news[sample(1:length(news),10000)]
sampleBlogs <- blogs[sample(1:length(blogs),10000)]
sampleData <- c(sampleTwitter,sampleNews,sampleBlogs)
writeLines(sampleData, ".sampleData.txt")

# remove temporary variables
rm(twitter,news,blogs,sampleTwitter,sampleNews,sampleBlogs,sampleData)
```

#Create and Clean Corpus

Using the tm package, the sampled data is used to create a corpus. Subsequently, the the following transformations are performed:

-convert to lowercase
-characters /, @ |
-common punctuation
-numbers
-English stop words
-strip whitespace
-stemming (Porter’s stemming)

```{r, echo = TRUE}
install.packages('tm', repos='http://cran.us.r-project.org')
library(tm)
cname <- file.path(".")
docs <- Corpus(DirSource(cname))

# convert to lowercase
docs <- tm_map(docs, content_transformer(tolower))

# remove more transforms
toSpace <- content_transformer(function(x, pattern) gsub(pattern, " ", x))
docs <- tm_map(docs, toSpace, "/|@|\\|")

# remove punctuation
docs <- tm_map(docs, removePunctuation)

# remove numbers
docs <- tm_map(docs, removeNumbers)

# strip whitespace
docs <- tm_map(docs, stripWhitespace)

# remove english stop words
docs <- tm_map(docs, removeWords, stopwords("english"))

# initiate stemming
install.packages('SnowballC', repos='http://cran.us.r-project.org')
library(SnowballC)
docs <- tm_map(docs, stemDocument)
```

#N-gram Tokenization

N-grams models are created to explore word frequencies. Using the RWeka package, unigrams, bigrams and trigrams are created.

```{r, echo = TRUE}
install.packages('RWeka', repos='http://cran.us.r-project.org')
library(RWeka)
Tokenizer <- function(x) NGramTokenizer(x, Weka_control(min = 1, max = 1))
unidtm <- DocumentTermMatrix(docs, control = list(tokenize = Tokenizer))

BigramTokenizer <- function(x) NGramTokenizer(x, Weka_control(min = 2, max = 2))
bidtm <- DocumentTermMatrix(docs, control = list(tokenize = BigramTokenizer))

TrigramTokenizer <- function(x) NGramTokenizer(x, Weka_control(min = 3, max = 3))
tridtm <- DocumentTermMatrix(docs, control = list(tokenize = TrigramTokenizer))
```

#Exploratory Data Analysis

##Top 10 Frequencies

Below, you can see the top 10 unigrams with the highest frequencies.

```{r, echo = TRUE}
tm_unifreq <- sort(colSums(as.matrix(unidtm)), decreasing=TRUE)
tm_uniwordfreq <- data.frame(word=names(tm_unifreq), freq=tm_unifreq)
paste("Unigrams - Top 5 highest frequencies")
```

```{r, echo = TRUE}
head(tm_uniwordfreq,5)
```

```{r, echo = TRUE}
tm_bifreq <- sort(colSums(as.matrix(bidtm)), decreasing=TRUE)
tm_biwordfreq <- data.frame(word=names(tm_bifreq), freq=tm_bifreq)
paste("Bigrams - Top 5 highest frequencies")
```

```{r, echo = TRUE}
head(tm_biwordfreq,5)
```

```{r, echo = TRUE}
tm_trifreq <- sort(colSums(as.matrix(tridtm)), decreasing=TRUE)
tm_triwordfreq <- data.frame(word=names(tm_trifreq), freq=tm_trifreq)
paste("Trigrams - Top 5 highest frequencies")
```

```{r, echo = TRUE}
head(tm_triwordfreq,5)
```

#Explore Frequencies

In the diagrams below, you can explore the Ngrams by frequencies:

```{r, echo = TRUE}
install.packages('ggplot2', repos='http://cran.us.r-project.org')
install.packages('dplyr', repos='http://cran.us.r-project.org')
library(ggplot2)
library(dplyr)
tm_uniwordfreq %>% 
    filter(freq > 1000) %>%
    ggplot(aes(word,freq)) +
    geom_bar(stat="identity") +
    ggtitle("Unigrams with frequencies > 1000") +
    xlab("Unigrams") + ylab("Frequency") +
    theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r, echo = TRUE}
tm_biwordfreq %>% 
    filter(freq > 100) %>%
    ggplot(aes(word,freq)) +
    geom_bar(stat="identity") +
    ggtitle("Bigrams with frequencies > 100") +
    xlab("Bigrams") + ylab("Frequency") +
    theme(axis.text.x=element_text(angle=45, hjust=1))
```

```{r, echo = TRUE}
tm_triwordfreq %>% 
    filter(freq > 10) %>%
    ggplot(aes(word,freq)) +
    geom_bar(stat="identity") +
    ggtitle("Trigrams with frequencies > 10") +
    xlab("Trigrams") + ylab("Frequency") +
    theme(axis.text.x=element_text(angle=45, hjust=1))
```

Below, we can see wordclouds of the top 50 unigrams, bigrams and trigrams.

##Wordcloud - Top 50 Unigrams

```{r, echo = TRUE}
install.packages('wordcloud', repos='http://cran.us.r-project.org')
library(wordcloud)
set.seed(39)
wordcloud(names(tm_unifreq), tm_unifreq, max.words=50, scale=c(5, .1), colors=brewer.pal(6, "Dark2"))
```

##Wordcloud - Top 50 Bigrams

```{r, echo = TRUE}
wordcloud(names(tm_bifreq), tm_bifreq, max.words=50, scale=c(5, .1), colors=brewer.pal(6, "Dark2"))
```

##Wordcloud - Top 50 Trigrams

```{r, echo = TRUE}
wordcloud(names(tm_trifreq), tm_trifreq, max.words=50, scale=c(5, .1), colors=brewer.pal(6, "Dark2"))
```
